---
title: "Train-Data-Interview"
author: "Anna Fetter"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)
library(ggrepel)
```
## Load Data 
```{r}
X22_blocked_crossings <- read_excel("data/2022_blocked_crossings.xlsx")
X23_blocked_crossings <- read_excel("data/2023_blocked_crossings.xlsx")
X24_blocked_crossings <- read_excel("data/2024_blocked_crossings.xlsx")

#join all 3 sets together, downloading more than a year at a time from the website was really finnicky

blocked_crossings <- bind_rows(X22_blocked_crossings, X23_blocked_crossings, X24_blocked_crossings)
```

## Taking the data "out for coffee"
```{r}
summary(blocked_crossings)
#unique(blocked_crossings$Reason)
#unique(blocked_crossings$`Immediate Impacts`)
#unique(blocked_crossings$Railroad)
#unique(blocked_crossings$Duration)
```
From initial data interview, it looks like there are problems with 2-6 hour delays being listed as "2-6 hours", "2-6 hours'", and "2-6 hours\"". I will change them all to "2-6 hours" for sake of clean and consistent data.
```{r}
blocked_crossings <- blocked_crossings %>%
  mutate(Duration = str_replace_all(Duration, "2-6 hours'", "2-6 hours")) %>%
  mutate(Duration = str_replace_all(Duration, "2-6 hours\"", "2-6 hours"))

unique(blocked_crossings$Duration)
```

Let's see how many states we have listed.
```{r}
blocked_crossings %>%
  group_by(State) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
```

Huh, only 49 states, let's match against an existing list of states and see who isn't on the list
```{r}
# built-in state abbreviations
state_abbreviations <- state.abb

# get distinct state abbreviations
observed_states <- blocked_crossings %>%
  distinct(State) %>%
  pull(State)

# identify unknown state abbreviations
invalid_states <- setdiff(observed_states, state_abbreviations)

# view them
print(invalid_states)

# let's see what states are missing that are in abbreviations but not observed states
missing_states <- setdiff(state_abbreviations, observed_states)

# view missing states
print(missing_states)
```

Based on this analysis DC is in the dataset, but Hawaii and Rhode Island are missing. A quick google search says that Hawaii has no active commercial railroads now. Rhode Island has some, but it's a super small state so not having complaints isn't a worry.

To make things easier later on, let's separate the year, month, and time into separate columns
```{r}
blocked_crossings <- blocked_crossings %>%
  mutate(Date = as.Date(`Date/Time`, format = "%m/%d/%Y")) %>%
  mutate(Year = year(Date)) %>%
  mutate(Month = month(Date)) %>%
  mutate(Time = format(as.POSIXct(`Date/Time`, format="%H:%M"), format="%H:%M"))

# as discovered in a later step, there is inconsistent capitalization, make everything uppercase for consistency across the dataset
blocked_crossings <- blocked_crossings %>% 
  mutate(City = str_to_upper(City)) %>%
  mutate(State = str_to_upper(State)) %>%
  mutate(Railroad = str_to_upper(Railroad)) %>%
  mutate(County = str_to_upper(County)) %>%
  mutate(`Crossing ID` = str_to_upper(`Crossing ID`)) %>%
  mutate(Reason = str_to_upper(Reason)) %>%
  mutate(`Immediate Impacts` = str_to_upper(`Immediate Impacts`)) %>%
  mutate(`Additional Comments` = str_to_upper(`Additional Comments`))
```

## 1.How many total complaints were filed in each year: 2022, 2023, and 2024?
```{r}
count(X22_blocked_crossings)
count(X23_blocked_crossings)
count(X24_blocked_crossings)

```
There were 30749 complaints in 2022, 19306 complaints in 2023, and 26687 complaints in 2024.

## 2. Which railroad company is associated with the most complaints?
```{r}
railroad_max_complaints <- blocked_crossings %>%
  group_by(Railroad) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(railroad_max_complaints)
```
Union Pacific, UP, had 28027 complaints between 2022-2024.

## 3.What are the top 10 states with the highest number of blocked train crossing complaints?
```{r}
states_most_blocked <- blocked_crossings %>% 
  group_by(State) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)

print(states_most_blocked)
```
Texas has the most blocked train crossings at 19,100. Illinois is second, with 6534 complaints over the three year period.

## 4. What are the top 10 cities with the highest number of blocked train crossing complaints?
```{r}
cities_most_blocked <- blocked_crossings %>% 
  group_by(City) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)

print(cities_most_blocked)
```
Houston has the most reported blocked crossings, at nearly 12000 reports within a three year span.


## 5. Which crossing IDs had the most complaints?
```{r}
# make sure city and state of crossing ID is included
crossings_most_blocked <- blocked_crossings %>%
  group_by(`Crossing ID`, City, State) %>%
  summarise(Count = n(), .groups = "drop") %>%
  #I had to drop the groups so they table would properly show up
  arrange(desc(Count)) %>%
  slice_head(n = 10)

print(crossings_most_blocked)
```
Houston, Texas has seven of the ten most reported blocked crossings in the country. Crossing ID 859522Y had 1388 reports in 3 years. 

(It could be interesting to dig more into this crossing in Houston for a story)

## 6. Which states have seen the greatest increase in complaints from 2022 to 2024?
```{r}
increase_in_complaints <- blocked_crossings %>%
  group_by(State, Year) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, values_from = Count) %>%
  mutate(Increase = `2024` - `2022`) %>%
  mutate(Percent_Increase = (Increase / `2022`) * 100) %>%
  arrange(desc(Increase))

print(increase_in_complaints)

#this is a bad metric, since the crossings with less complaints that got more with time will show super high percent increases
percent_increase_in_complaints <- increase_in_complaints %>% 
  arrange(desc(Percent_Increase))

print(percent_increase_in_complaints)
```
Texas has seen the greatest total increase in blocked train crossing complaints from 2022 to 2024, with 1,762 more complaints (+27.1%). It’s followed by Virginia, which saw an increase of 960 complaints (+183.2%), and California with 474 more complaints (+81.9%) over the same period.

## 7. What is the distribution of blocked train crossing delays? What is the quantity/percentage over 15 minutes?
```{r}
delay_distribution <- blocked_crossings %>% 
  group_by(Duration) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100) %>%
  arrange(desc(Count))

ggplot(delay_distribution, aes(x = Duration, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Blocked Train Crossing Delays (2022-2024)",
       x = "Duration",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

delays_over_fifteen <- blocked_crossings %>% 
  filter(Duration != "0-15 minutes")

percent_delays_over_fifteen <- nrow(delays_over_fifteen)/nrow(blocked_crossings) * 100

print(percent_delays_over_fifteen)
```
89% of reported delays were over 15 minutes.

## 8. What is the distribution of complaints by day of the week?
```{r}
dow_complaints <- blocked_crossings %>% 
  mutate(Day_of_Week = weekdays(Date)) %>%
  group_by(Day_of_Week) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

ggplot(dow_complaints, aes(x = Day_of_Week, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Complaints by Day of the Week (2022-2024)",
       x = "Day of the Week",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## 9. How many blocked crossings were reported during peak commute times?
```{r}
#let's do this by mapping what the distribution is for every single hour during the day
blocked_crossings <- blocked_crossings %>%
  mutate(Hour = hour(`Date/Time`)) %>% 
  mutate(Peak_Commute = ifelse(Hour >= 7 & Hour <= 9 | Hour >= 16 & Hour <= 18, "Yes", "No"))

ggplot(blocked_crossings, aes(x = Hour)) +
  geom_bar(aes(fill = Peak_Commute), position = "dodge") +
  labs(title = "Blocked Crossings During Peak Commute Times (2022-2024)",
       x = "Hour of the Day",
       y = "Count") +
  scale_fill_manual(values = c("Yes" = "hotpink", "No" = "grey")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```
## 10. What day had the most train crossing complaints nationwide?
```{r}
# find what day had the most complaints nationwide
day_most_complaints <- blocked_crossings %>%
  group_by(Date) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 1)

print(day_most_complaints)
```
Nationwide May 20, 2022 saw the most blocked crossing complaints at 243 total complaints. 

## 11. What specific crossing had the most complaints in a single day?
```{r}
# find what specific crossing had the most complaints in a single day
top_complaint_day_crossing <- blocked_crossings %>% 
  group_by(`Crossing ID`, Date, City, State) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))

print(top_complaint_day_crossing) 


# based on the results from top_complaint_day_crossing, let's look to see what's the distribution of the number of complaints, maybe a box and whiskers plot
ggplot(top_complaint_day_crossing, aes(x = Count)) +
  geom_boxplot() +
  labs(title = "Distribution of Complaints per CrossingID",
       x = "Crossing ID",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# lowkey a useless graph, except for showing that most crossing complaints were like 1 per incident
  
  
```
 Crossing ID 859523F in Houston, TX saw the most complaints in a single day at a single crossing at 98 total complaints.

## 12. What is the split between complaints of moving, stationary trains, or no train was present but the gate was activated?
```{r}
#asked chatgpt for styling help, hence ggrepel
train_status <- blocked_crossings %>%
  count(Reason) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label = paste0(Reason, "\n", n, " (", percent, "%)"),
    ypos = cumsum(n) - 0.5 * n  # position for label
  ) 

plot_train_status <- train_status %>%
  ggplot(aes(x = "", y = n, fill = Reason)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text_repel(
    aes(y = ypos, label = label),
    nudge_x = 1.3,
    show.legend = FALSE,
    segment.size = 0.3,
    size = 3
  ) +
  labs(
    title = "Train Status Complaints (2022–2024)",
    fill = "Reason"
  ) +
  theme_void() +
  theme(legend.position = "none")  # << this hides the sidebar legend

print(plot_train_status)

```

## 13. How many complaints list no train present but still report gates down or signals activated?
```{r}
train_status %>% 
  filter(str_detect(Reason, "NO TRAIN WAS PRESENT")) %>%
  arrange(desc(n))
```
2284 complaints listed the reason for blockage as "no train was present but the lights and/or gates were activated".

## 14. How many complaints mentioned “first responders were observed being unable to cross the tracks?”
```{r}
responders_blocked <- blocked_crossings %>% 
  filter(str_detect(`Immediate Impacts`, "FIRST RESPONDERS"))

unique(blocked_crossings$`Immediate Impacts`)

print(responders_blocked)

nrow(responders_blocked)
```
Over 15000 complaints mentioned that first responders were blocked by the train crossing.

## 15. How many complaints mentioned “pedestrians were observed climbing on, over, or through train cars?”
```{r}
pedestrians_crossings <- blocked_crossings %>% 
  filter(str_detect(`Immediate Impacts`, "PEDESTRIANS"))

print(pedestrians_crossings)

#percentage of citings that mention pedestrians
percent_pedestrians <- nrow(pedestrians_crossings)/nrow(blocked_crossings) * 100

print(percent_pedestrians)
```
Almost 18,000 complaints, 23% of all complaints, cited pedestrians climbing over, under, or through the blocked train crossing.

## 16. What percentage of comments cited “traffic”?
```{r}
#using string detect method I learned in PLAN372
mentioned_traffic <- blocked_crossings %>% 
  filter(str_detect(`Additional Comments`, regex("Traffic", ignore_case = TRUE)))

print(mentioned_traffic)

percent_mentioned_traffic <- nrow(mentioned_traffic)/nrow(blocked_crossings) * 100

print(percent_mentioned_traffic)
```
Nearly 10% of complaints explicitly mentioned "traffic" in their additional comments about the incident, totalling over 7,500 complaints from 2022-2024.

## 17. Which incidents received the most complaints? Which date/time and crossing ID combination appears most frequently in the dataset?
```{r}
incidents_most_complaints <- blocked_crossings %>%
  mutate(rounded_time = floor_date(`Date/Time`, unit = "hour")) %>%
  group_by(rounded_time, `Crossing ID`, City, State, Railroad) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)

incidents_most_complaints

## I NEED TO DOUBLE CHECK THIS, is date and time too rigid, should it be a range?

## THIS FEELS APPROXIMATE, I would not use this in an actual story, more as a way to find interesting blockings & use as anecdotes
```
See comment above. Would not include this analysis in a news story since it doesn't feel precise enough.

## 18. How many complaints were filed per railroad company
```{r}
complaints_per_railroad_company <- blocked_crossings %>% 
  group_by(Railroad) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

complaints_per_railroad_company

# identify top 5 railroads with most complaints
top_railroads <- complaints_per_railroad_company %>%
  slice_head(n = 5) %>%
  pull(Railroad)

#asked chatgpt with help making pie chart
blocked_crossings_pie <- blocked_crossings %>%
  mutate(railroad_group = ifelse(Railroad %in% top_railroads,
                                 Railroad, "Other Railroads")) %>%
  count(railroad_group, name = "Count") %>%
  arrange(desc(Count)) %>%
  mutate(perc = round(Count / sum(Count) * 100, 1),
         label = paste0(railroad_group, " (", perc, "%)"))

ggplot(blocked_crossings_pie, aes(x = "", y = Count, fill = railroad_group)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  guides(fill = guide_legend(title = "Railroad")) +
  geom_text(aes(label = ifelse(Count > 0, label, "")),
            position = position_stack(vjust = 0.5),
            size = 3)
```

## 19. What is the average complaint duration by company?
```{r}
# I need to double check that since duration is a character string, it's in the appropriate order
  # answer:chatgpt suggested using ordered factors to make sure durations are in correct order
duration_factors <- c("0-15 minutes",
                     "16-30 minutes",
                     "1-2 hours",
                     "2-6 hours",
                     "6-12 hours",
                     "12-24 hours",
                     "More than one day")

# keeping it to the top 5 so I know the sample is appropriately large
duration_per_company <- blocked_crossings %>%
  mutate(Duration = factor(Duration, levels = duration_factors, ordered = TRUE)) %>% 
  filter(Railroad %in% top_railroads) %>%
  group_by(Railroad, Duration) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))

duration_per_company

# I need to double check that since duration is a character string, it's in the appropriate order
  # answer:chatgpt suggested using ordered factors to make sure durations are in correct order
unique(blocked_crossings$Duration)

duration_factors <- c("0-15 minutes",
                     "16-30 minutes",
                     "1-2 hours",
                     "2-6 hours",
                     "6-12 hours",
                     "12-24 hours",
                     "More than one day")


# make sure to use median, since I'm working with ranges
# Median duration as a category (not numeric)
median_duration_per_company <- blocked_crossings %>%
  filter(Railroad %in% top_railroads, !is.na(Duration)) %>%
  group_by(Railroad, Duration) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(Railroad, Duration) %>%
  group_by(Railroad) %>%
  mutate(CumSum = cumsum(Count), Total = sum(Count)) %>%
  filter(CumSum >= Total / 2) %>%
  slice_head(n = 1) %>%
  select(Railroad, Median_Duration = Duration)

median_duration_per_company

#huh, returning all 16-30 minutes, let's make bar charts for all the 5 to see average delay
ggplot(duration_per_company, aes(x = Railroad, y = Count, fill = Duration)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Complaint Duration by Railroad Company (2022-2024)",
       x = "Railroad Company",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The top 5 railroad companies by complaint volume had an average reported blocked crossing time between 16-30 minutes.

## 20. Which counties had the highest number of complaints across all three years?
```{r}
complaint_counties <- blocked_crossings %>% 
  group_by(County, State) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)

print(complaint_counties)

#oh no! there are different reports based on capitalization! I need to go back through and clean up blocked crossings so that it doesn't care about capitalization
```
Harris County, Texas received the most complaints between 2022-2024, totalling over 12,800 reports of blocked crossings. Cook County, IL lagged far behind in second place with 3723 reported blocked crossings.

## 21. How many complaints were filed in each month across the dataset?
```{r}
blocked_crossings <- blocked_crossings %>%
  mutate(Month = factor(month.name[as.numeric(Month)], levels = month.name, ordered = TRUE))

complaints_per_month <- blocked_crossings %>% 
  group_by(Month) %>% 
  summarise(Count = n(), .groups = "drop") %>%
  arrange(Month)

print(complaints_per_month)

ggplot(complaints_per_month, aes(x = Month, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Blocked Train Crossing Complaints by Month (2022–2024)",
       x = "Month",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mean(complaints_per_month$Count)
```
August had the most complaints per month from 2022-2024, totalling nearly 7400. The mean complaints per month from that period is just under 6400.

## 22. How many total crossing locations appear only once in the dataset?
```{r}
one_hit_wonders <- blocked_crossings %>% 
  group_by(`Crossing ID`, City, State) %>%
  summarise(Count = n(), .groups = "drop") %>%
  filter(Count == 1)

one_hit_wonders

nrow(one_hit_wonders)/nrow(blocked_crossings)
```
About 10% of crossings in the dataset received only one complaint in the three year span.


# North Carolina specific questions
```{r}
# start by filtering the data for just NC
nc_blocked_crossings <- blocked_crossings %>% 
  filter(State == "NC")

nc_blocked_crossings
```

## 23. How many complaints were filed in North Carolina in total from 2022 to 2024?
```{r}
nc_total_complaints <- nrow(nc_blocked_crossings)

print(nc_total_complaints)
```
There were 411 complaints of blocked train crossings in North Carolina between 2022-2024.

## 24. Which city in North Carolina had the most blocked train crossing complaints?
```{r}
nc_city_top_blockings <- nc_blocked_crossings %>% 
  group_by(City) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

nc_city_top_blockings
```
Rocky Mount, NC had the most reporting blocked train crossings, at 176 in the 3 year span.

## 25. What county in North Carolina had the most blocked train crossing complaints?
```{r}
nc_county_top_blockings <- nc_blocked_crossings %>% 
  group_by(County, City) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

nc_county_top_blockings
```
Nash County, home to Rocky Mount, had the most reported blocked crossings.

## 26. What are the most frequently reported railroad companies in North Carolina complaints?
```{r}
nc_top_company_blockings <- nc_blocked_crossings %>% 
  group_by(Railroad) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(nc_top_company_blockings)
#I wonder what the total percent of complaints CSX had, seems about 70% of all NC complaints if I had to guess
```

## 27. What percent of total complaints in NC was CSX responsible for?
```{r}
csx_blocked_percent_nc <- nc_top_company_blockings %>% 
  filter(Railroad == "CSX") %>%
  summarise(Percent = (Count / nc_total_complaints) * 100)

print(csx_blocked_percent_nc)
```
CSX had 282 reported complaints from 2022-2024, accounting for 69% of all complaints in North Carolina over the three year span.

## 28. Which crossing IDs in North Carolina were reported the most frequently?
```{r}
# based on other questions asked earlier, this might be redundant, I'm going to guess it's in Rocky Mount
nc_crossings_most_blocked <- nc_blocked_crossings %>%
  group_by(`Crossing ID`, City, County) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))

print(nc_crossings_most_blocked)
```
A crossing in Rocky Mount, NC received the most complaints-- 167 in a 3 year period.

## 29. Follow up question, when were all the Rocky Mount, NC complaints, was it a one time event or does it happen routinely?
```{r}
rocky_mount_blocked <- nc_blocked_crossings %>% 
  group_by(`Crossing ID`, Date, Railroad) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))

rocky_mount_blocked

# looks like it was dominated by a few huge delay days, all operated by CSX
```
Rocky Mount's complaints were accumulated most on 4 distinct dates: May 20, 2024 (49 complaints), January 3, 2024 (19 complaints), September 18, 2024 (16 complaints), and September 14, 2024 (11 complaints).

## 30. What percentage of North Carolina complaints involved blockages longer than 15 minutes?
```{r}
more_than_fifteen_minutes_nc <- nc_blocked_crossings %>% 
  filter(Duration != "0-15 minutes")

percent_more_than_fifteen_nc <- nrow(more_than_fifteen_minutes_nc)/nrow(nc_blocked_crossings) * 100

percent_more_than_fifteen_nc
```

87% of reported complaints in NC were complaints with reported delay times of more than 15 minutes.

## 31. How many North Carolina complaints mentioned “first responders were observed being unable to cross the tracks?”
```{r}
nc_responders_blocked <- nc_blocked_crossings %>% 
  # this string detect only works as well as it does like this bc this outputs came from a form, fixed word choice
  filter(str_detect(`Immediate Impacts`, "FIRST RESPONDERS"))

nrow(nc_responders_blocked)
```
63 complaints mentioned first responders were observed unable to cross the tracks.

## 32. What county had the most complaints that mentioned “first responders were observed being unable to cross the tracks?”
```{r}
nc_responders_blocked_county <- nc_responders_blocked %>% 
  group_by(County) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(nc_responders_blocked_county)
```
Nash County, home of Rocky Mount, had 21 reports of first responders unable to cross the tracks due to a blocked train crossing.

## 33. How many North Carolina complaints mentioned “pedestrians were observed climbing on, over, or through train cars?”
```{r}
nc_pedestrians_blocked <- nc_blocked_crossings %>% 
  filter(str_detect(`Immediate Impacts`, "PEDESTRIANS"))

nc_pedestrians_blocked
```
55 complaints mentioned “pedestrians were observed climbing on, over, or through train cars."
## 34. What county had the most complaints that mentioned  “pedestrians were observed climbing on, over, or through train cars?”
```{r}
nc_county_pedestrians_blocked <-nc_pedestrians_blocked %>% 
  group_by(County) %>% 
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(nc_county_pedestrians_blocked)
```
Of those 55 complaints, 25 were reported in Mecklenburg county and 15 were reported in Nash County.

** Based on this analysis, my story pitch will likely involve digging deeper into those blocked crossings in Rocky Mount.**